# Creating Graphics with `ggplot2`

The package [ggplot2](http://ggplot2.tidyverse.org/) provides an abstract and declarative environment for creating graphics.

The graphics system built into R is already quite powerful and flexible, but creating sophisticated graphics can be time-consuming and many steps that could be performed automatically, like adding a legend, have to be performed manually. Code producing more complex visualisations tends be "procedural": rather than describing how the visualisation should look like, the code describes the detailed control flow of how the plot is constructed.


In this section, we will use the `health` dataset. You can download this data to your R console by running the following command:

```{r}
load(url("https://github.com/UofGAnalyticsData/R/raw/main/Week%205/w5.RData"))
```

## ggplot terms 

This section gives an overview of key terms in the `ggplot2` world. `ggplot2` is based on the philosophy of a "layered grammar of graphics": plots in `ggplot2` are made up of at least one layer of geometric objects. 


### Geometric objects 
A geometric object (or `geom_<type>(...)` in `ggplot2` commands) controls what type of plot a layer contains. The are many different geometric objects: the most important ones are ...

| Geometry name | Description | Basic R equivalent| Common aesthetics |
|---------------|--------------------------|----------------|------------------|
|`geom_point`| Points (scatter plot) | `plot` / `points` | `x`, `y`, `alpha`, `colour`, `shape`,`size`|
|`geom_line`| Lines (drawn left to right)  | `lines` (after ordering) | `x`, `y`, `alpha`, `colour`,`linetype`, `size` |
| `geom_path` | Lines (drawn in original order) | `lines` | `x`, `y`, `alpha`, `colour`, `group`,`linetype`, `size` |
|`geom_abline` | Line (one line) | `abline`  | `intersept`, `slope`, `alpha`,`colour`, `linetype`, `size`|
| `geom_hline`| Horizontal line | `abline` | `yintercept`, `alpha`,`colour`, `linetype`, `size`|
| `geom_vline`| Vertical line | `abline` | `xintercept`, `alpha`, `colour`,`linetype`, `size`|
| `geom_text` | Text | `text` | `x`, `y`, `label`, `alpha`, `angle`, `colour`, `size`, `family`, `hjust`,`vjust`, `check_overlap` | 
| `geom_label`| Text (styled as label) | `text` | `x`, `y`, `label`, `alpha`, `angle`, `colour`, `size`, `family`, `hjust`, `vjust`, `check_overlap`|
| `geom_rect` | Rectangle | `rect`| `xmin`, `xmax`, `ymin`, `ymax`,`alpha`, `colour`, `fill`,`linetype`, `size` | 
| `geom_polygon`| Polygon | `polygon`| `x`, `y`, `alpha`, `colour`, `fill`,`group`, `linetype`, `size`|
| `geom_ribbon` | Ribbon (for confidence bands) | - | `x`, `ymin`, `ymax`, `alpha`, `colour`,`fill`, `group`, `linetype`,`size` |
| `geom_bar`| Bar plot | `barplot`| `x`, `alpha`, `colour`, `fill`,`linetype`, `size`|
| `geom_boxplot`| Boxplot | `boxplot` |  `x`, `y`, `alpha`, `colour`, `fill`,`group`, `linetype`,`shape`, `size` | 
| `geom_histogram`| Histogram | `hist` | `x`, `y`, `alpha`, `colour`, `fill`,`linetype`, `size` |
| `geom_raster` /`geom_tile`  | Image plot  | `image` | `x`, `y`, `alpha`, `fill` (both) and `linetype`, `size`, `width` (`geom_tiles` only) |
| `geom_counter`| Contour lines | `contour` | `x`, `y`, `z`, `alpha`, `colour`, `group`,`linetype`,`size` |


There is a [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) providing a detailed overview of the different geometries and data.

### Aesthetics 

An aesthetic (or `aes(...)` in `ggplot2` commands) controls which variables are mapped to which properties of the geometric objects (like x-coordinates, y-coordinates, colours, etc.). The aesthetics available depend on the geometric object. Aesthetics commonly available are ...

| Aesthetic     | Description                       |
|--------|-------------------------------|
| `x`    | x-coordinate   |
| `y`    | y-coordinate   |
| `color` or `colour` | Colour (outline) |
| `fill`   | Fill colour  |
| `alpha`  | Transparency (transparent $0\leq \alpha\leq 1$ opaque)| 
| `linetype` | Line type ("`lty`")|
| `symbol`  | Plotting symbol ("`pch`") |
| `size`   | Size of plotting symbol / font or line thickness | 

The help file for each geometry lists the available aesthetics.


## Using the more general ggplot interface

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
load(url("https://github.com/UofGAnalyticsData/R/raw/main/Week%205/w5.RData"))
```

### A typical ggplot call

A plotting command for `ggplot` consists of a sequence of function calls added together using the standard sum operator `+`:
```r
ggplot(data=...) +                # Specify data source
    aes(...) +                    # Generic aesthetics applying to all layers
    geom_<type>(aes(...), ...)  + # Geometry for one layer with layers-specific aesthetics
    geom_<type>(aes(...), ...)  +
    ...                           # Further arguments for fine-tuning (themes, scales, facets, ...)
```

`geom_<type>` objects do not necessarily have to use the same data as specified in the call to `ggplot`. If the optional argument `data` is specified, then the data source provided is used for this layer.

For the following example, we will use data on health expenditure by country, reported on an annual basis. We can produce a plot like the following using `ggplot` commands. 

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
ggplot(data=health) + 
  aes(x=HealthExpenditure,  y=LifeExpectancy) +
  geom_point(aes(colour=Region, size=Population)) +
  scale_x_log10()
```

### Adding additional layers

Additional layers can simply be added to the plot. For example, we can add an overall regression line with confidence bands using

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy) +
    geom_point(aes(colour=Region, size=Population)) +
    geom_smooth(method="lm")  +
    scale_x_log10()
```

If we want to add a different regression line for each country we have to make sure that a `group` or `colour` aesthetic is passed to `geom_smooth`. We could pass `aes(colour=Region)` to `geom_smooth`. Alternatively, we can move `colour=Region` from the aesthetics specific to `geom_point` to the generic aesthetics, so that `colour=Region` now applies to both  `geom_point` and `geom_smooth`.

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy, colour=Region) +
    geom_point(aes(size=Population)) +
    geom_smooth(method="lm")  +
    scale_x_log10()
```

The warning comes from the fact that there are only two North American countries, so we can fit a line through them with no error, which means we cannot draw confidence bands. 

The plot looks slightly messy, we will use `facet_wrap` later on to split it into separate panels.

Suppose we want to annotate the observations belonging to Australia, the UK, the US.
```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
health2 <- health %>%
    filter(Country %in% c("Australia", "United Kingdom", "United States"))
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy, colour=Region) +
    geom_point(aes(size=Population)) +
    geom_label(data=health2, 
               aes(x=HealthExpenditure,  y=LifeExpectancy, label=Country),
               show.legend=FALSE)
```

The labels however cover the observations and might not be fully visible. This can be avoided by using the function `geom_label_repel` from [ggrepel](https://cran.r-project.org/web/packages/ggrepel).
```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
health <- health %>%
    mutate(CountryLabel=ifelse(Country%in%c("Australia", "United Kingdom", "United States"),
                               as.character(Country),""))

library(ggrepel)
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy, colour=Region) +
    geom_point(aes(size=Population)) +
    geom_label_repel(aes(label=CountryLabel), show.legend=FALSE)
```

This time, we have used a different approach. Rather than subsetting the data and creating a separate data frame only containing the data for the three countries, we have created a new column in the data frame `health`, which is blank except for the three countries. This is required because `ggrepel` layers are only aware of data drawn in their own layer: this way we can avoid the labels covering observations we have not labelled.



### Explicit drawing
The standard R plotting functions draw a plot as soon as the plot function is invoked. 

Plotting commands in `ggplot2` (including `qplot`) return objects (otherwise the `+` notation would not work) and only draw the plot when their `print` or `plot`  methods are invoked. In the console this is the case when they are used without an assignment. 

```r
a <- ggplot(data=health) +            # Does not draw anything
         aes(x=HealthExpenditure, y=LifeExpectancy) +
         geom_point()

b <- a + scale_x_log10()              # Does not draw anything either

a                                     # Now the plot stored in a gets drawn
print(a)                              # Draw a again (explicit invocation)

b                                     # Now the plot stored in b gets drawn
```

Inside loops and functions the `print` or `plot` methods need to be invoked explicitly by using the methods `print` or `plot`. 

### Task
Consider two vectors `x` and `y` created using
```r
n <- 1e3
x <- runif(n, 0, 2*pi)                   # x is random uniform from (0,2*pi)
# x <- sort(x)                           # Sorting of x _not_ needed for ggplot
y <- sin(x)                              # Set y to the sine of x
y.noisy <- y + .25 * rnorm(n)            # Create noisy version of y
```
Use `ggplot2` to create a scatterplot of `y.noisy` against `x`, which also shows the noise-free sine curve in `y`.

### Answer
We can use the following R code:
```{r, echo=FALSE}
n <- 1e3
x <- runif(n, 0, 2*pi)                   # x is random uniform from (0,2*pi)
y <- sin(x)                              # Set y to the sine of x
y.noisy <- y + .25 * rnorm(n)            # Create noisy version of y
```
```{r, dev.args=list(pointsize=10)}
ggplot() +                               # No need to use data=... as x, y and y.noisy
                                         # are variables in the workspace and not columns
                                         # in a dataset
    geom_point(aes(x, y.noisy)) +
    geom_line(aes(x, y))
```

It does not matter whether `geom_point` or `geom_line` comes first. `ggplot2` adapts the axes so that all objects drawn fit (and not just the first one as is the case when using standard R plotting functions `plot` and `points`).

## Modifying Plots

### Labels and titles

We can set the plot title using `ggtitle(title)` and the axis labels using `xlab(label)` and  `ylab(label)`.

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth", warning=FALSE}
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy, colour=Region) +
    geom_point(aes(size=Population)) +
    geom_smooth(method="lm")  +
    scale_x_log10() +
    ggtitle("Relationship between Health Expenditure and Life Expectancy") +
    xlab("Health Expenditure") +
    ylab("Life Expectancy")    
```

Changing the text shown in legends (like  in our case the names of the regions) is more complicated. It is almost always easier to simply change the levels of the categorical variable in the dataset itself before invoking `ggplot2` commands.


### Scales
Aesthetics control *which* variables are mapped to *which* property of the geometric object. However, aesthetics do not specify *how* this mapping is performed. This is where scales come into play. Scales control *how* any value from the variable is translated into a property of a geometric object: scales control for example how a variable is translated into coordinates (say through a log transform) or into colours (say though a discrete colour palette).

`ggplot2` automatically chooses (what it thinks is) a suitable scale. This is usually reasonable, but on occasions it might be necessary to override this.

There is a family of scale functions for each aesthetic. The template for the function name for scales is `scale_<aesthetic>_<type>`. 

#### Scales for continuous data

We have already seen that we can log-transform the axes using `scale_x_log10` and `scale_x_log10`. The more general functions for coordinate transforms are `scale_<x or y>_continous(...)`. We can. amongst others,  set the axis label (argument `name`, the ticks and tickmarks (arguments `breaks` and `labels`) the limits (argument `limit`) and the transform to be used (argument `trans`).

The axes might use scientific notation (e.g. "4e5"). If you want to avoid using scientific notation and use fixed notation, change the `scipen` option in R, which controls when scientific notation is used (for example run `options(scipen=1e3)`).

There are functions for mapping continuous data to other aesthetics, too. For example, `scale_colour_gradient` converts a continuous variable to a colour using a gradient of colours. The arguments `low` and `high` specify the colours used at the two ends. `scale_colour_gradient2` allows for also specifying a mid-point colour (argument `mid`). `scale_colour_gradientn` is the most general function it allows specifying a vector of colours and corresponding vector of colours. The function `scale_colour_distiller` uses the colour brewer available at [http://colorbrewer2.org/](colorbrewer.org) and allows for constructing colours scales which are photocopier-safe and/or work for colour-blind readers. 

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
a <- ggplot(data=health) +
         aes(x=HealthExpenditure, y=LifeExpectancy) +
         geom_point(aes(colour=Population)) +
         scale_colour_distiller(palette="YlOrRd" , trans="log")
a
```

We have used `trans="log"` to use the log-transformed values of the population sizes (due to its skewness). The values given in the legend seem slightly odd choices: this is due to the log-transform (they are roughly $\exp(14)$, $\exp(17)$ and $\exp(20)$, so "nice" numbers on the log scale). 

We have stored the plot in a variable `a` so that we can redraw it later on with different themes.



#### Scales for discrete data

There are also various scaling functions for discrete data, such as `scale_colour_brewer`. 

Note that there are separate scales for colour (outline colour -- example: `scale_colour_brewer`) and fill (fill colour -- example: `scale_fill_brewer`).


### Statistics
Sometimes data has to be aggregated before it can be used in a plot. For example, when creating a bar plot illustrating the distribution of a categorical variable we have to count how many observations there are in each category. This will then determine the height of the bars. `ggplot2` automatically chooses (what it thinks is) a suitable statistic. 

For example, when we draw a bar plot using `geom_bar`, it uses by default the statistic `count`, which first produces a tally. We don't need to worry about this, `ggplot2` does all the work for us.
```{r}
ggplot(data=health) +
    geom_bar(aes(x=Region)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x axis labels
```

Suppose we now want to a draw bar chart visualising the mean health expenditure in each region. Now we don't want `ggplot2` to produce a tally of how often which value occurs, we want it to simply draw the bars to the heights specified in the data. Because we now want no aggregation, we have to use the statistic `identity`.
```{r}
library(dplyr)
HESummary <- health %>%                                      # Get avg health exp
                 group_by(Region) %>% 
                 summarise(HealthExpenditure=mean(HealthExpenditure))
ggplot(data=HESummary) +
    geom_bar(aes(x=Region, y=HealthExpenditure), stat="identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x axis labels
```

### Theming
Themes can be used to customise how `ggplot2` graphics look like. We have already used `theme` to change how the horizontal axis is typeset. 

`ggplot2` has several themes built-in. The default theme is `theme_gray`. Other themes available are `theme_bw` (monochrome), `theme_light`, `theme_lindedraw` and `theme_minimal`. Further themes are available in extension packages such [ggthemes](https://cran.r-project.org/web/packages/ggthemes).

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
a + theme_minimal()
library(ggthemes)
a + theme_economist() + theme(legend.position="right")
```

### Arranging plots (faceting)

The function `facet_grid(rvar~cvar)` creates separate plots based on the values `rvar` (rows) and `cvar` (columns) takes. The function `facet_wrap(~var1+var2)` arranges the plots in several rows and columns without rigidly associating one variable with rows and one with columns. Continuous variables need to be discretised (for example using `cut`) before they can be used for defining facets.

```{r, dev.args=list(pointsize=10), fig.width=10, fig.height=6, out.width="\\textwidth"}
ggplot(data=health) + 
    aes(x=HealthExpenditure,  y=LifeExpectancy, colour=Region) +
    geom_point(aes(size=Population)) +
    geom_smooth(method="lm")  +
    scale_x_log10() +
    facet_wrap(~Region)
```

Arranging plots in more general ways (like in `par(mfrow=c(...))` or `layout`) is not directly possible with `ggplot2`. The package [gridExtra](https://cran.r-project.org/web/packages/gridExtra) however provides a function `grid.arrange`, which allows for arranging `ggplot2` plots side by side.


